Performance analysis of high-resolution \\ ice sheet simulations

### 2022-10-09 paper/reviews/response.tex

\maketitle

%\tableofcontents

\thispagestyle{empty}
%\bigskip

The thoughtful comments of the two reviewers are much appreciated.  Below I both respond to their concerns---in block italics---and, in the important cases, quote the new or rewritten content of the paper which addresses these concerns.  The aggregate responses to these comments have made the paper about one page longer, i.e.~when printed in the final two-column format.

\section{Responses to Scientific Editor Ralf Greve}

\begin{review}
As for the first paragraph of the introduction and the comment by Reviewer 2 on L25, it may also be worth mentioning the findings by Gladstone et al. (2017, TC, doi: 10.5194/tc-11-319-2017). They showed that the resolution requirement can be alleviated substantially by choosing proper representations for basal sliding and melting that ensure continuity across the grounding line.
\end{review}

\noindent Yes, that makes sense.  This idea and citation is integrated into my response to Reviewer 2 on L25.

@@

Unfortunately ``Asymptotic analysis of implicit versus explicit methods in ice sheet models'' would not clarify which is the asymptotic limit.   (It is not aspect ratio $H/L \to 0$, nor approach to steady state $t\to \infty$, nor even convergence under the high-resolution $\Delta x \to 0$, but rather performance in the number of degrees of freedom limit.)  My intent is to address \emph{performance}, as a broader issue which relates to multiple modeling choices, specifically mesh resolution, time-stepping type (explicit vs.~implicit), solver design (multigrid versus not), and even operational ice-climate coupling frequency, all over long-duration runs.

In other words, I don't see how to write a title of acceptable length without requiring the interested reader to (at least) dive into the abstract for clarification of some brief phrase such as ``performance analysis''.  (A shorter title may itself increase the number of such interested readers.)  A possible title is ``Performance analysis of numerical ice sheet simulations in the high-resolution limit'', and this identifies the asymptotic, but I find the current title to be essentially as clear, and clearly briefer.  Does the Editor have a title suggestion?

\begin{review}
\textbf{L25} 'it is generally accepted...'  I am not sure that I agree that this is generally accepted.  There remain important works in Antarctica where models are run at 20km resolution or more and I’m not sure that they are ‘invalid’.  It is worth perhaps being more specific about the resolution requirements of specific tasks, e.g. capturing detailed perturbations to the grounding line or calving front, the influence of steep subglacial topography, etc.
\end{review}

\noindent I stand by the assertion, but this reasonable challenge has caused me to add some words to clarify.  The sentence in question is now expanded into two sentences which explain some of the needs for resolution:

\begin{quote}``In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows and alpine topography need yet finer resolution.  Resolving the physics of marine ice sheets also benefits from fine resolution, although careful choice of basal parameterizations will reduce resolution dependence (Gladstone and others, 2017).''\end{quote}

\noindent A few points about the first of these new sentences will, I think, be understood by any reader with minimal exposure to ice sheet modeling:
\begin{itemize}
\item This is a necessarily-imprecise description of practical resolution needs for ``resolving'' physical mechanisms understood to be important by a community of scientists.  This is not a mathematically- or statistically-based assertion.
\item The added phrase ''ice streams as fluid features'' refers to aspects of model results which are necessarily more than one grid cell wide.  That is, as fluids are locally infinite-dimensional, you need a few cells, more than one or two across for ice streams, to capture a ``fluid feature''.
\item 10 km resolution is itself not enough for yet narrower or smaller features.
\item The added word ``it is \underline{now} generally accepted'' clearly refers to expectations in the 2020s.  There were indeed ``important'' 20km results before now, but I would disagree with any assertion in a present-day journal submission that, for example, outlet glaciers in Antarctica or Greenland are ``validly modeled'' at 20 km resolution; that would now be a consequence of not trying very hard, or choosing a deficient model.
\end{itemize}
With that background, I am happy if readers conclude, for example, that the author thinks that 20km or 40km ice sheet model results are no longer ``valid'' in a scientific sense.  That is what I think, and I was under the impression that this was an acceptable judgement call in the community.

### 2022-10-09 paper/perfmod.tex
ready to ship out

\abstract{Numerical ice sheet models compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.  At high spatial resolution, with horizontal mesh/grid resolutions of a few kilometers or smaller, these models usually require time steps shorter than climate-coupling time scales because they update ice thickness after each velocity solution.  High-resolution performance is degraded by the stability restrictions of such explicit time-stepping.  This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, clarifies the scaling of numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.  The performance of current-generation explicit time-stepping models is assessed, and then compared to the prospective performance of implicit schemes.  The main result, Table \ref{tab:performancemodel}, highlights the key roles played by the algorithmic scaling of stress-balance and coupled, implicit-step solvers.}

@@

Numerical ice sheet (glacier) models with evolving ice geometry are now routinely used for addressing scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenlandic \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.  In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows need yet finer resolution.  Whether using local mesh refinement \citep[for example]{Fischleretal2022,Hoffmanetal2018} or not, resolutions of two or one kilometers \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019,Clarkeetal2015} are increasingly used for science at ice sheet scale.

@@

For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged horizontal velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, the mass continuity equation says
\begin{equation}
\frac{\partial H}{\partial t} + \Divx \left(\bu H\right) = a, \label{eq:masscontinuity}
\end{equation}
where $\bx=(x,y)$ denotes the horizontal coordinates.  Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.  This is because ice flows dominantly downhill.  Indeed, ice sheet flow has no characteristic curves, as would Equation \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.

Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of  advective and diffusive mechanisms.  Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  A numerical thickness perturbation will often cause the velocity $\bu$ to respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term effectively acts along the surface gradient $\gradx s$ because of the ice-domain geometry.  One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\gradx s)$, a non-local function of geometry; a stress balance solution is required to evaluate this function.  A numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step.

@@

However, either statement \eqref{eq:masscontinuity} or \eqref{eq:siamasscontinuity} of the mass continuity equation is fundamentally incomplete without modeling the changing (map-plane) boundary position of the glacier or ice sheet.  The problem of determining this updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).  The mathematical role of this constraint has been understood for some time in the SIA model \citep{Calvoetal2002,JouvetBueler2012,SchoofHewitt2013}, but its role as the determinant of margin location is universal across fluid-layer problems with signed source terms \citep{Bueler2021conservation}.

@@

Assuming the map-plane extent of the model domain is 2D and of width $L$, these primary parameters are related by
\begin{equation}
\Delta x \sim \frac{L}{m^{1/2}} \quad \text{and} \quad m \sim \frac{L^2}{\Delta x^2}.  \label{eq:paramrelation}
\end{equation}
Specifically, for fixed domain width $L$ there are $O(m^{1/2})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$, but this paper addresses 3D models with map-plane (2D) horizontal meshes.}

@@

Glacier and ice sheet models resolve and integrate ice-climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and on longer scales.  Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.  Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.  Note that energy balance and degree-day schemes for computing surface mass balance \citep{GreveBlatter2009} generally have much shorter time scales, but here $q$ describes the \emph{coupling} frequency on which ice geometry needs to be updated via solution of the mass continuity or surface kinematical equation.

@@

Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$, but this restriction reflects the simulation purpose, not maintenance of stability.  For an implicit scheme the desired frequency of ice-climate interaction will, however, determine the total simulation cost according to the (large) solution cost, at each time step, of solving coupled mass and momentum equations, including the surface kinematical or mass continuity equation.

@@

When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and it depends strongly on solver design.  Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling, as actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.  Many models \citep[e.g.][]{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} also show good \emph{parallel} scaling, something not addressed here.

@@

Now, how many flops are needed to simulate one model year?  Suppose a numerical model takes time steps of $\Delta t$ model years duration, equivalently $\Delta t^{-1}$ steps per model year.  In these terms stability restrictions give required numbers of steps per model year.  That is, stability requires that $\Delta t^{-1}$ be bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.  Recalling the scaling in Equation \eqref{eq:paramrelation}, the explicit SIA and pessimistic-Stokes cases have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}
The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$ for Stokes models, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

@@

% regarding next paragraph, see data/bueler.perf and data/scaling.m
For SIA models, let us assume that the flops of such coupled solutions scale as $O(m^{1+\beta})$ with $\beta \ge 0$; a large constant is assumed.  The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.  For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, directly computes steady states with scaling $\beta=0.8$.  (An earlier steady, Picard iteration implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is also robust for implicit time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic and irregular bed topographies.

@@

Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of membrane-stress-resolving solvers dominates performance concerns.  In particular, an $\alpha=1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition.  Optimistic stability yields a still-bad $O(\Delta x^{-5})$ rate.  These simple observations emphasize the key role which will be played by multigrid-based stress balance solvers.  That is, even retaining explicit time-stepping, nearly-optimal ($\alpha \approx 0$) scaling of computation effort in Stokes and higher-order solvers will be necessary for routine application on high-resolution meshes.

On the other hand, suppose resolution ($\Delta x$) is fixed.  The Table also shows why algorithmic scaling remains important in the large-domain $L\to\infty$ limit when applying Stokes or higher-order dynamics.  The computational work of an optimal solver will be proportional to ice sheet area $L^2$.  However, an $\alpha=1$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.

@@

>>>>

\abstract{Numerical glacier and ice sheet models compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.  At high spatial resolution, with horizontal mesh/grid resolutions of a few kilometers or smaller, these models usually require time steps shorter than climate-coupling time scales because they update ice thickness after each velocity solution.  High-resolution performance is degraded by the stability restrictions of such explicit time-stepping.  This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, clarifies the scaling of numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.  The performance of current-generation explicit time-stepping models is assessed, and then compared to the prospective performance of implicit schemes.  The main result, Table \ref{tab:performancemodel}, highlights the key roles played by the algorithmic scaling of stress-balance and coupled, implicit-step solvers.}

@@

Numerical ice sheet (glacier) models with evolving ice geometry are now routinely used for addressing scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenlandic \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.  In order to resolve ice streams as fluid features it is now generally accepted that valid results need horizontal mesh (grid) cells smaller than about 10 km, but narrow outlet glacier flows and alpine topography need yet finer resolution.  Resolving the physics of marine ice sheets also benefits from fine resolution, although careful choice of basal parameterizations will reduce resolution dependence \citep{Gladstoneetal2017}.  Thus, whether using local mesh refinement \citep[for example]{Fischleretal2022,Hoffmanetal2018} or not, resolutions of two or one kilometers \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019,Clarkeetal2015} are increasingly used for science at ice sheet scale.

@@

For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged horizontal velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, the mass continuity equation says
\begin{equation}
\frac{\partial H}{\partial t} + \Divx \left(\bu H\right) = a. \label{eq:masscontinuity}
\end{equation}
(Here $\bx=(x,y)$ denotes the horizontal coordinates.)  Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.  This is because ice flows dominantly downhill.  Indeed, ice sheet flow has no characteristic curves, as would Equation \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.

Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of  advective and diffusive mechanisms.  Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  A numerical thickness perturbation will often cause the velocity $\bu$ to respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term effectively acts along the surface gradient $\gradx s$ because of the ice-domain geometry.  One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\gradx s)$, a non-local function of geometry, but a stress balance solution is required to evaluate this function.  In any case, a numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step \citep{LeVeque2007}.

@@

However, either statement \eqref{eq:masscontinuity} or \eqref{eq:siamasscontinuity} of the mass continuity equation is fundamentally incomplete without modeling the evolving (map-plane) boundary position of the glacier or ice sheet.  The problem of determining this updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).  The mathematical role of this constraint has been understood for some time in the SIA model \citep{Calvoetal2002,JouvetBueler2012,SchoofHewitt2013}, but its role as the determinant of margin location is universal across fluid-layer problems with signed source terms \citep{Bueler2021conservation}.

@@

Assuming the map-plane extent of the model domain is 2D\footnote{In flow-line models $\Delta x \sim L m^{-1}$, but this paper addresses 3D models with map-plane (2D) horizontal meshes.} and of width $L$, these primary parameters are related by
\begin{equation}
\Delta x \sim \frac{L}{m^{1/2}} \quad \text{and} \quad m \sim \frac{L^2}{\Delta x^2}.  \label{eq:paramrelation}
\end{equation}
Specifically, for fixed domain width $L$ there are $O(m^{1/2})$ mesh cells in each horizontal dimension.

@@

Glacier and ice sheet models resolve and integrate ice-climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and on longer scales.  Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.  Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.  Note that energy balance and degree-day schemes for computing surface mass balance \citep{Hock2005} generally have much shorter time scales, but here $q$ describes the \emph{coupling} frequency on which ice geometry needs to be updated via solution of the mass continuity or surface kinematical equation.

@@

Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$, but this restriction reflects the simulation purpose, not maintenance of stability.  For an implicit scheme the desired frequency of ice-climate interaction will, however, determine the total simulation cost according to the (large) solution cost, at each time step, of solving a coupled mass and momentum free-boundary problem, including the surface kinematical or mass continuity equation.

@@

When analyzing solver scaling in our simplified form, one must be aware that the constant in ``$O(m^{1+\alpha})$'' above can be very large, and it depends strongly on solver design.  Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling, as actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.  Indeed, many models \citep[e.g.][]{BrownSmithAhmadia2013, Fischleretal2022, IsaacStadlerGhattas2015, Lengetal2012, Tuminaroetal2016} also show good \emph{parallel} scaling, something not addressed here.

@@

Now, how many flops are needed to simulate one model year?  Suppose a numerical model takes time steps of $\Delta t$ model years duration, equivalently $\Delta t^{-1}$ steps per model year.  In these terms stability restrictions give required numbers of steps per model year.  That is, stability requires that $\Delta t^{-1}$ be bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.  Recalling the scaling in Equation \eqref{eq:paramrelation}, the explicit SIA and pessimistic-Stokes cases have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}
The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$ for Stokes models and $O(m)$ for SIA models, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

@@

% regarding next paragraph, see data/bueler.perf and data/scaling.m
For SIA models, let us assume that the flops of such coupled solutions scale as $O(m^{1+\beta})$ with $\beta \ge 0$; a large constant is assumed.  The only currently implemented, unconditionally-stable, fully-implicit time-step solvers use the SIA stress balance.  For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, directly computes steady states with scaling $\beta=0.8$.  (An earlier steady, Picard iteration implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is also robust for implicit time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic and irregular bed topographies.

@@

Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of membrane-stress-resolving solvers dominates performance concerns.  In particular, an $\alpha \approx 1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition.  Optimistic stability yields a still-bad $O(\Delta x^{-5})$ rate.  These simple observations emphasize the key role which will be played by multigrid-based stress balance solvers.  That is, even retaining explicit time-stepping, nearly-optimal ($\alpha \approx 0$) scaling of computation effort in Stokes and higher-order solvers will be necessary for routine application on high-resolution meshes.

On the other hand, suppose resolution ($\Delta x$) is fixed.  The Table also shows why algorithmic scaling remains important in the large-domain $L\to\infty$ limit when applying Stokes or higher-order dynamics.  The computational work of a nearly-optimal $\alpha \approx 0$ solver will be proportional to ice sheet area $L^2$.  However, an $\alpha \approx 1$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha \approx 4$) power of $L$.

@@

\subsection{Acknowledgements}  Comments by Jed Brown, editor Ralf Greve, referee Alex Jarosch, and especially an anonymous referee have helped to focus and improve this work.  However, opinions expressed here, and any remaining errors, are entirely those of the author.
